---
title: "01 Data Downloads"
output: html_document
author: ML Gaynor
date: "2025-07-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose: 
- Download and preview species occurrence records using iDigBio and gatoRs

Load Required Packages
```{r}
library(ridigbio)     # Interface to iDigBio API
library(gatoRs)       # Unified taxonomic/occurrence tools
library(leaflet)      # Interactive mapping
```

#### Retrieving data from iDigBio using ridigbio
Search the iDigBio API for occurrence record by taxonomic designations
```{r}
# Search for specific species (Galax urceolata)
iDigBio_GU <- idig_search_records(rq = list(scientificname = "Galax urceolata"))
# Search for all Diapensiaceae occurrences (limit to 1000 for example)
iDigBio_GU_family <- idig_search_records(rq = list(family = "Diapensiaceae"), limit = 1000)
```
Search the API with a geographic bounding box (e.g. Eastern USA extent)
```{r}
# format a request in the structure of a list
rq_input <- list(
  scientificname = list(type = "exists"),
  family = "Diapensiaceae",
  geopoint = list(
    type = "geo_bounding_box",
    top_left = list(lon = -98.16, lat = 48.92),
    bottom_right = list(lon = -64.02, lat = 23.06)
  )
)
# input request into idigbio search fxn
iDigBio_GU_family_USA <- idig_search_records(rq_input, limit = 1000)
```

Save results as a Comma Separated File (.csv)
```{r}
write.csv(iDigBio_GU, "data/01_download/iDigBio_GU_2025_06_27.csv", row.names = FALSE)
write.csv(iDigBio_GU_family, "data/01_download/iDigBio_GU_family_2025_06_27.csv", row.names = FALSE)
```

#### Retrieving Data from iDigBio and GBIF using gatoRs
Create a synonym vector for species to search by. 
```{r}
Shortia_galacifolia <- c("Shortia galacifolia", "Sherwoodia galacifolia")
Galax_urceolata <- c("Galax urceolata", "Galax aphylla")
Pyxidanthera_barbulata <- c("Pyxidanthera barbulata", "Pyxidanthera barbulata var. barbulata")
Pyxidanthera_brevifolia <- c("Pyxidanthera brevifolia", "Pyxidanthera barbulata var. brevifolia")

```
Use gatoRs to download records species occurrence data for each name vector. Allows us to retrieve all versions of a name as taxonomy has been updated. 
```{r eval=FALSE, include=TRUE}
gators_download(synonyms.list = Shortia_galacifolia,
                write.file = TRUE,
                filename = "./data/01_download/raw/Shortia_galacifolia_raw_2025_06_27.csv")

gators_download(synonyms.list = Galax_urceolata,
                write.file = TRUE,
                filename = "./data/01_download/raw/Galax_urceolata_raw_2025_06_27.csv")

gators_download(synonyms.list = Pyxidanthera_barbulata,
                write.file = TRUE,
                filename = "./data/01_download/raw/Pyxidanthera_barbulata_raw_2025_06_27.csv")

gators_download(synonyms.list = Pyxidanthera_brevifolia,
                write.file = TRUE,
                filename = "./data/01_download/raw/Pyxidanthera_brevifolia_raw_2025_06_27.csv")

```

#### Preview a sample species's occurrence data
Read in Shortia galacifolia's data we downloaded with gatoRs above
```{r}
rawdf <- read.csv("data/01_download/raw/Shortia_galacifolia_raw_2025_06_27.csv")
```
View number of columns/fields in the dataset, also return the number of rows (individual occurrence records)
```{r}
names(rawdf) # returns field headers
nrow(rawdf) # returns the number of rows = occurrence records
```

Use Leaflet to visualize records interactively - The error message here indicates many points do not have long/lat values 
```{r}
leaflet(rawdf) %>% 
  addTiles() %>% 
  addMarkers(label = paste0(rawdf$longitude, ", ", rawdf$latitude))
```

