---
title: "02 Data Cleaning"
output: html_document
date: "2025-07-21"
Author: ML Gaynor
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose: 
- Clean and harmonize occurrence records for ecological niche modeling.

Load Required Packages
```{r message=FALSE}
library(gatoRs)         # Taxonomic and geographic cleaning
library(fields)         # For calculating spatial distances
library(sf)             # For spatial operations and visualization
library(ggplot2)        # For plotting
library(ggspatial)      # Scale bars and north arrows
library(leaflet)        # For interactive mapping
```

#### Read in Raw Occurrence Data
Read in raw species-specific data downloaded in 01 Data Downloads. Check how many occurrence records we start with in the raw data. 
```{r}
rawdf <- read.csv("data/01_download/raw/Shortia_galacifolia_raw_2025_06_27.csv")
nrow(rawdf)  # Starting number of records
```

#### Taxonomic Harmonization 
Our initial download includes iDigBio & GBIF's synonym lists in addition to the name vectors we searched. Let's inspect and clean to our taxonomic standard. Note that taxa_clean() contains the argument taxa.filter which allow the options "exact", "fuzzy" or "interactive" to allow us to tighten or loosen the requirements for name matching. Read more by running ?taxa_clean
```{r}
# Inspect unique names
unique(rawdf$scientificName)
# Define our synonym list
search <- c("Shortia galacifolia", "Sherwoodia galacifolia")
# Clean names with fuzzy matching
df <- taxa_clean(df = rawdf,
                 synonyms.list = search,
                 taxa.filter = "fuzzy", # fuzzy allows some leeway over exact matching
                 accepted.name = "Shortia galacifolia")
nrow(df)  # Updated count
```
#### Locality Cleaning 
Here we remove any records with missing coordinates, impossible coordinates, coordinates at (0,0), and any that are flagged as skewed. We also round the provided latitude and longitude values to a specified number of decimal places.
```{r}
# Remove invalid/skewed records and round coordinates
df <- basic_locality_clean(df = df,
                           remove.zero = TRUE,
                           precision = TRUE,
                           digits = 2,
                           remove.skewed = TRUE)
nrow(df)

```

#### Flag and Filter Cultivated Records 
Remove coordinates in cultivated zones, botanical gardens, and outside our desired range. Here we set interactive = FALSE, however you can visualize the outliers by setting interactive = TRUE. 
```{r}
df <- process_flagged(df, interactive = FALSE, scientific.name = "accepted_name")
nrow(df)
```

#### Remove Duplicates
Here we identify and remove both (1) specimen duplicates and (2) aggregator duplicates based on each specimens coordinates, occurrenceID, and eventDate. To leverage all date information available, set remove.unparseable = FALSE to manually populate the year, month, and day columns. Here, we also confirm all ID (UUID and key) are unique to remove any within-aggregator duplicates that may accumulate due to processing errors.
```{r}
# Remove duplicate specimens and aggregator artifacts
df <- remove_duplicates(df, remove.unparseable = TRUE)
nrow(df)
```

#### Spatial Deduplication
Maxent will only retain one point per pixel. To make the ecological niche analysis comparable, we will retain only one point per pixel. Note: Default is a raster with 30 arc sec resolution. 
```{r}
# Retain one point per raster pixel (30 arc-sec default)
df <- one_point_per_pixel(df)
nrow(df)
```

#### Spatial Thinning
Oversampled regions/habitats can bias our model. To reduce bias, we can thin occurrence data. <br>

First, lets assess the current minimum distance among your occurrence records. Here, we calculate the minimum nearest neighbor distance in km.
```{r}
# Calculate minimum nearest-neighbor distance
nnDm <- rdist.earth(as.matrix(df[, c("longitude", "latitude")]), miles = FALSE)
nnDmin <- do.call(rbind, lapply(1:5, function(i) sort(nnDm[, i])[2]))
min(nnDmin)  # e.g., 2.22 km
```
Here the current minimum distance is 2.22 km. Based on literature, we find a 2 meters (or 0.002 km) distance was enough to collect unique genets, so we do not need to thin our points. <br>

However when you do need to thin your records, this is a great function in gatoRs to do so! 
```{r}
# thin if you need
df <- thin_points(df, 
                    distance = 0.002, # in km 
                    reps = 100)
```

#### Creating a Static Map of Cleaned Occurrences 
Make points spatial 
```{r}
df_fixed <- st_as_sf(df, coords = c("longitude", "latitude"), crs = 4326)
```

Set up base maps 
```{r}
USA <- borders(database = "usa", colour = "gray80", fill = "gray80")
state <- borders(database = "state", colour = "black", fill = NA)
```

Plot: Use ggplot2 & ggspatial to plot occurrence records. We are using two base maps, USA and State. The geom_sf() fxn adds our cleaned occurrence points as blue based on their lon & lat values. We can also fix coordinate spatial view through coord_sf(). Finally can add a scale and arrow. 
```{r}
simple_map <- ggplot() +
  USA + state +
  geom_sf(data = df_fixed, color = "blue") +
  coord_sf(xlim = c(min(df$longitude) - 3, max(df$longitude) + 3),
           ylim = c(min(df$latitude) - 3, max(df$latitude) + 3)) +
  xlab("Longitude") + ylab("Latitude") +
  annotation_scale(plot_unit = "km") +
  annotation_north_arrow(location = "tl", height = unit(1, "cm"), width = unit(1, "cm"))

simple_map

```
#### Interactive Mapping
Another fun way to visualize occurrence data is using leaflet's interactive maps. 
```{r}
leaflet(df_fixed) %>%
  addMarkers(label = paste0(df$longitude, ", ", df$latitude)) %>%
  addTiles()
```
#### Write out the cleaned occurrence data as a .csv file
```{r}
write.csv(df, "data/02_cleaning/Shortia_galacifolia_2025_06_27_cleaned.csv", row.names = FALSE)
```

#### Batch clean all of the species 
Use a for-loop to iterate through all of the species. 
```{r}
files <- list.files("data/01_download/raw", full.names = TRUE)[1:3]
synonymns <- list(
  Galax_urceolata = c("Galax urceolata", "Galax urceolata (Poir.) Brummitt", "Galax urceolata (Poiret) Brummitt", "Galax urceolaa", "Galax aphylla L.", "Galax aphylla"),
  Pyxidanthera_barbulata = c("Pyxidanthera barbulata", "Pyxidanthera barbulata Michx.", "Pyxidanthera barbulata var. barbulata", "Pyxidenthera barbulata"),
  Pyxidanthera_brevifolia = c("Pyxidanthera brevifolia", "Pyxidanthera brevifolia Wells", "Pyxidanthera barbulata var. brevifolia (Wells) H.E.Ahles", "Pyxidanthera barbulata var. brevifolia")
)

for (i in 1:3) { # for loop just allows us to iterate through each species
  df <- read.csv(files[3])
  search <- synonymns[[3]]

  df <- taxa_clean(df, synonyms.list = search, taxa.filter = "fuzzy", accepted.name = search[1])
  df <- basic_locality_clean(df, remove.zero = TRUE, precision = TRUE, digits = 2, remove.skewed = TRUE)
  df <- process_flagged(df, interactive = FALSE, scientific.name = "accepted_name") # removing all points for brevifolia
  df <- remove_duplicates(df, remove.unparseable = TRUE)
  df <- one_point_per_pixel(df)
  df <- thin_points(df, distance = 0.002, reps = 100)

  outfile <- paste0("data/02_cleaning/", gsub(" ", "_", search[1]), "_2025_06_27_cleaned.csv")
  write.csv(df, outfile, row.names = FALSE)
  rm(df, search, outfile)
}

```
#### Merge Cleaned Files for Maxent
```{r}
alldf <- list.files("data/02_cleaning", pattern = "*.csv", full.names = TRUE)
alldf <- lapply(alldf, read.csv)
alldf <- do.call(rbind, alldf)
write.csv(alldf, "data/02_cleaning/maxent_ready/diapensiaceae_maxentready_2025_06_27.csv", row.names = FALSE)

```

#### Map all records 
```{r}
alldf_fixed <- st_as_sf(alldf, coords = c("longitude", "latitude"), crs = 4326)

all_map <- ggplot() +
  USA + state +
  geom_sf(data = alldf_fixed, aes(col = factor(accepted_name))) +
  coord_sf(xlim = c(min(alldf$longitude) - 3, max(alldf$longitude) + 3),
           ylim = c(min(alldf$latitude) - 3, max(alldf$latitude) + 3)) +
  xlab("Longitude") + ylab("Latitude") +
  labs(color = "Scientific name") +
  annotation_scale(plot_unit = "km") +
  annotation_north_arrow(location = "tl", height = unit(1, "cm"), width = unit(1, "cm"))

all_map
```

#### Prep Geolocate Batch file
Some records do not contain lat/lon info, but do have useful information in the locality field that can give us a reasonable estimate of location + some error. Here we'll use the gatoRs fxn need_to_georeference() to isolate these records for batch processing on the georeferencing platform GeoLocate. <br> 

Read in _raw_ data and check what needs to be georeferenced
```{r}
rawdf <- read.csv("data/01_download/raw/Shortia_galacifolia_raw_2025_06_27.csv")
rawdf_GeoRef <- need_to_georeference(rawdf)
```
Format the columns/fields for GeoLocate's standards, then write out for batch processing. 
```{r}
# Format columns for GeoLocate submission
rawdf_GeoRef <- rawdf_GeoRef %>%
  dplyr::select(
    "locality string" = locality,
    country,
    state = stateProvince,
    county,
    latitude,
    longitude,
    ID,
    name = scientificName,
    basis = basisOfRecord
  )

# Add required empty columns
rawdf_GeoRef$'correction status' <- ""
rawdf_GeoRef$precision <- ""
rawdf_GeoRef$'error polygon' <- ""
rawdf_GeoRef$'multiple results' <- ""

# Reorder columns
rawdf_GeoRef2 <- rawdf_GeoRef[, c("locality string", "country", "state", "county", "latitude", "longitude",
                                  "correction status", "precision", "error polygon", "multiple results",
                                  "ID", "name", "basis")]

# write out csv
write.csv(rawdf_GeoRef2, "data/03_georeferencing/Shortia_galacifolia_Needing_GeoRef_2025_06_27.csv", row.names = FALSE)

```




