---
title: "03 Data Exploration"
output: html_document
date: "2025-07-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose: 
- Analyze climatic variables associated with species occurrences using PCA and ANOVA.
- plot top contributors to PC1 and PC2.

Load Packages
```{r message=FALSE}
library(gtools)         # For mixedsort of file names
library(terra)          # For raster and spatial extraction
library(dplyr)          # For data wrangling
library(tidyr)          # For handling NAs and reshaping
library(ggplot2)        # For plotting
library(ggbiplot)       # For PCA biplot visualization
library(multcompView)   # For compact letter display (Tukey HSD)
library(gridExtra)      # For combining plots
```

#### Load in Cleaned Occurrence Data
```{r}
# Read in cleaned occurrence data with geographic coordinates and species ID
alldf <- read.csv("data/02_cleaning/maxent_ready/diapensiaceae_maxentready_2025_06_27.csv")
```

#### Load in and stack bioclim layers
Raster files provided by WorldClim 
```{r}
# List and sort all .tif files in the WorldClim directory
list <- list.files("data/04_climate_processing/BioClim/", full.names = TRUE)
list <- gtools::mixedsort(sort(list))

# Stack the raster files
envtStack <- terra::rast(list)
```

#### Extract Climatic Variables at Occurrence Points

Extract climatic variable from raster cell where occurrence pts sit 
```{r}
# Extract values of environmental rasters at occurrence points
ptExtracted <- terra::extract(envtStack, alldf[, c("longitude", "latitude")])
```

Organize these data to have species information
```{r}
# Combine with species names and coordinates
ptExtracteddf <- ptExtracted %>%
  dplyr::mutate(name = as.character(alldf$accepted_name), 
                x = alldf$longitude, 
                y = alldf$latitude)

```

Drop any points that does not contain climatic data
```{r}
# Remove any rows with missing climate values
ptExtracteddf <- ptExtracteddf %>% tidyr::drop_na()
```

#### Perform PCA on Climatic Variables
Subset the climatic variables, assumed to be columns 2 to 21
```{r}
data.bioclim <- ptExtracteddf[, 2:21]
data.species <- ptExtracteddf[, "name"]
```

Scale and run a Principal Component Analysis (PCA) on climatic variables. A PCA reduces the dimensionality of the climatic data to make it easier to identify broad patterns and which variables are most influential. For our data, it shows how species cluster & differentiate according to the climatic variables associated with their spatial presence. <br>

When you use the command prcomp your loading variables show up as rotational variables. 
```{r}
pca_result <- prcomp(data.bioclim, scale. = TRUE)
```

Extract variable loadings: saving a matrix of ceofficients (called loadings) for each of the original climatic variable on each PC (principal component). These loadings show how much each original variable contributes to each PC. Variables with larger (pos or neg) loadings are the biggest drivers of variation in the data. 
```{r}
loadings <- pca_result$rotation
```
Calculate relative contributions of each variable to each PC
```{r}
loadings_relative <- sweep(abs(loadings), 2, colSums(abs(loadings)), "/") * 100
```
#### Visualize the PCA
Set up a nice custom theme for plotting aesthetics 
```{r}
# Customize theme
custom_theme <- theme(
  panel.background = element_blank(),
  panel.border = element_rect(fill = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  strip.background = element_blank(),
  axis.ticks = element_line(colour = "black"),
  plot.margin = unit(c(1, 1, 1, 1), "line"),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12),
  legend.title = element_text(size = 12),
  text = element_text(size = 12)
)
# Define color palette for species
pal <- c("#D43F3AFF", "#EEA236FF", "#5CB85CFF", "#46B8DAFF")

```
Create a species biplot. obs.scale indicates the scale factor to apply to observation, var.sacle indicates the scale factor to apply to variables, ellipse as TRUE draws a normal data ellipse for each group, and circle as TRUE draws a correlation circle. <br>

This is all basically to show how our data clusters across the top 2 PCs per species!
```{r}
# PCA Biplot
pca_plot <- ggbiplot(pca_result, obs.scale = 1, var.scale = 1,
                     groups = data.species, ellipse = TRUE, circle = TRUE) +
  scale_color_manual(name = '', values = pal) +
  theme(legend.direction = 'vertical', legend.position = 'bottom',
        legend.text = element_text(size = 12, face = "italic")) +
  custom_theme

print(pca_plot)
```

#### Identify the top contributors to PC1 and PC2
Pull out the bioclim layers that comprise PC1 and PC2
```{r}
# Top 2 contributing variables to PC1 and PC2
top_PC1_vars <- rownames(loadings)[order(abs(loadings[, "PC1"]), decreasing = TRUE)][1:2]
top_PC2_vars <- rownames(loadings)[order(abs(loadings[, "PC2"]), decreasing = TRUE)][1:2]
selected_vars <- unique(c(top_PC1_vars, top_PC2_vars))  # remove duplicates
```

#### ANOVA + Tukey HSD + Plotting for Top Variables
We can run an ANOVA (Analysis of Variance) and a Tukey HSD (a comparison test) Post-hoc to test whether there are statistically significant differences in the average values of specific climatic variables between the species. <br> 

Broken down, the ANOVA evaluates the difference of the mean value for a climatic variable across all species, and tests for a significant difference. <br>

The Tukey HSD test follows up comparing all pairs of species to see which species pairs differ from eachother. <br>

Here, we'll use a for-loop to move through each extracted (via species occurrence points) bioclim layer for PC1 & PC2 and conduct the ANOVA and Tukey HSD. Plotted as a boxplot with real data to show relative density.  
```{r}
## ANOVA + Tukey HSD + Plotting for Top Variables ----

plotlist <- list()

for (i in seq_along(selected_vars)) {
  varname <- selected_vars[i]
  
  # Run ANOVA
  model <- aov(as.formula(paste0(varname, " ~ name")), data = ptExtracteddf)
  tukey <- TukeyHSD(model)
  pvals <- tukey$`name`[, 4]
  
  # Convert p-values to significance letters
  letters_out <- multcompLetters(pvals)
  groups_df <- data.frame(
    name = names(letters_out$Letters),
    groups = letters_out$Letters
  )
  
  # Set y position for labels
  max_y <- max(ptExtracteddf[[varname]], na.rm = TRUE)
  groups_df$y_position <- max_y * 1.1
  
  # Create plot
  plotlist[[i]] <- ggplot(ptExtracteddf, aes(x = name, y = .data[[varname]])) +
    geom_jitter(aes(color = name), width = 0.3, alpha = 0.5, size = 1) +
    geom_boxplot(fill = NA, color = "black", size = 0.6) +
    geom_text(data = groups_df,
              aes(x = name, y = y_position, label = groups),
              inherit.aes = FALSE, size = 4) +
    scale_fill_manual(values = pal) +
    scale_color_manual(values = pal) +
    ggtitle(paste0(varname)) +
    ylab(varname) +
    theme_classic() +
    theme(legend.position = "none", 
          axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")) +
    ylim(0, max_y * 1.3)
}

# Combine plots
gridExtra::grid.arrange(grobs = plotlist, ncol = 2)

```

